\documentclass[10pt,letterpaper]{article}
\usepackage[letterpaper,total={5.0in,9in}]{geometry}
% This is basically plain TeX, but I'm putting a LaTeX wrapper on it...
%
% Original is at:
% https://stackoverflow.com/questions/1732348/regex-match-open-tags-except-xhtml-self-contained-tags/1732454#1732454
\font\lil=cmr5%
\font\bfit=cmbxti10%
\def\rayz#1#2{\llap{\raise #1\hbox{#2}}}%
\def\rayzl#1#2{\llap{\raise #1\hbox{\lil #2}}}%
\def\rayzb#1#2{\llap{\raise #1\hbox{\pmb{#2}}}}%
\def\strike#1{\rlap{\raise 0.7ex\hbox{\vrule height 0.5pt width #1}}}%
\def\graybox#1{\rlap{\special{color push rgb 0.8 0.8 0.8}%
\lower 2pt\hbox{\vrule height 2ex width #1}\special{color pop}}}
\def\pmb#1{\setbox0=\hbox{#1}%
\kern-.01em\copy0\kern-\wd0%
\kern.05em\copy0\kern-\wd0%
\kern-.01em\raise.0333em\box0}%
\title{You Can't Parse HTML with Regex}
\author{Amusing StackOverflow Answer}
\date{circa 2009}

\begin{document}
\parskip 10pt plus 4pt minus 4pt\parindent=0pt%
\lineskiplimit=-1000pt% don't expand lines that are too large..
\maketitle

\parskip=0pt
You can't parse [X]HTML with regex. Because HTML can't be parsed by
regex. Regex is not a tool that can be used to correctly parse
HTML\hbox{}. As I have answered in HTML-and-regex questions here so many
times before, the use of regex will not allow you to consume
HTML\hbox{}. Regular expressions are a tool that is insufficiently
sophisticated to understand the constructs employed by HTML\hbox{}. HTML is
not a regular language and hence cannot be parsed by regular
expressions. Regex queries are not equipped to break down HTML into
its meaningful parts. so many times but it is not getting to me. Even
enhanced irregular regular expressions as used by Perl are not up to
the task of parsing HTML\hbox{}. You will never make me crack. HTML is a
language of sufficient complexity that it cannot be parsed by regular
expressions. Even Jon Skeet cannot parse HTML using regular
expressions. Every time you attempt to parse HTML with regular
expressions, the unholy child weeps the blood of virgins, and Russian
hackers pwn your webapp. Parsing HTML with regex summons tainted souls
into the realm of the living. HTML and regex go together like love,
marriage, and ritual infanticide. The $<$center$>$ cannot hold it is too
late. The force of regex and HTML together in the same conceptual
space will destroy your mind like so much watery putty. If you parse
HTML with regex you are giving in to Them and their blasphemous ways
which doom us all to inhuman toil for the One whose Name cannot be
expressed in the Basic Multilingual Plane, he comes. HTML-plus-regexp
will liquify the nerves of the sentient whilst you observe, your
psyche withering in the onslaught of horror.
Rege\rayz{0pt}{$\bar{\phantom{e}}$}%
\rayz{1pt}{$\bar{\phantom{e}}$}%
\rayz{2pt}{$\grave{\phantom{e}}$}%
\rayz{3pt}{$\acute{\phantom{e}}$}x-based
HTML parsers are the cancer that is killing StackOverflow
{\it it is too late it is too
late we cannot be saved} the transgression of a
chil\rayz{6.5pt}{$\frown$}d
ensures regex
will consume all living tissue (except for HTML which it cannot, as
previously prophesied)
{\it dear lord help us how can anyone survive this
scourge} using regex to parse HTML has doomed humanity to an eternity
of dread torture and security holes {\it using rege}x as a tool to process
HTML establishes a brea{\it ch between this world} and the dread realm of
$\dot{\rm c}$\rayz{1pt}{$\hat{\phantom{c}}$}\rayzl{8pt}{h}%
o\rayz{-1pt}{$^{{}^\varsigma}$}\rayzl{8pt}{m}%
rrupt
entities (like SGML entities, but
{\it more corrupt) a mere glimp}se of the world

\raggedright{}of reg{\bf ex parsers for HTML will ins}tantly transport a
p{\it ro\-grammer's consciousness i}nto a w{\it orl}d of ceaseless screaming, he
comes\strike{7em}, the pestilent slithy
regex-infection wil{\bf l devour your HT}ML
parser, application and existence for all time like Visual Basic only
worse {\it he comes he comes do not fi}ght
h{\bf e co\rlap{\lower 1pt\hbox{\ \ \thinspace,}}mes, hi}s
un\rayz{-1.35ex}{`}h$\overline{\rm ol}$y radia\'nc\'e
de{\it stroyi\rayz{-1.5ex}{$\hat{}$\ }%
\rayz{0pt}{`\thinspace$\check{}$\ }%
\rayzl{0pt}{>\ \thinspace}%
\rayzl{3pt}{<}%
ng all
enli\rayz{-1pt}{$\skew6\ddot{\phantom{i}}$}%
\rayz{0pt}{$\skew6\acute{\phantom{i}}$}%
\rayz{1.5pt}{$\skew6\hat{\phantom{i}}$}%
\rayz{3pt}{$\skew6\ddot{\phantom{i}}$}%
\rayz{4.5pt}{$\skew6\acute{\phantom{i}}$}%
ghtenment, HTML tags 
{\bfit leak\rayz{0pt}{$\widetilde{\phantom{ak}}$}\c{i}n\.g
fro\rayz{-1.3ex}{`\thinspace}\rayz{0pt}{\vrule height 0.7pt width 0.5em\ }m
you\rayz{-2pt}{$\underline{\phantom{ou}}$}r eye\rayz{-1.8ex}{$\vec{\phantom{e}}$}%
s\kern-4pt /\thinspace{}l\rayz{1.6ex}{,\hskip 1pt}i\rayz{2.5pt}{'\hskip 1pt}ke
liq}uid p}ain, the song of re\rlap{\kern-2pt/}gular
expre\strike{6em}ssion parsing will
exti{\it nguish
the voices of mor{\bfit tal man from the sp}here
I can see it can you see
$\hat\imath$\rayz{2pt}{$\acute{}\,$}%
\rayz{-2.5ex}{$\grave{}$\thinspace}%
\rayzl{-1.9ex}{<\thinspace}%
\rayz{-2.0ex}{$_{{}_\vdash}$}%
t\rayz{1pt}{$\acute{}$}\rayz{3pt}{$\skew8\acute{}\acute{\thinspace}$}\rayz{4.5pt}{$\skew7\grave{}$}%
\rayzl{0pt}{$_{{}_{{}_\infty}}$}%
\rayzl{-1.0ex}{t}%
\rayzl{-1.7ex}{l\thinspace}%
\rayz{-2.5ex}{--}%
\rayzl{-2.5ex}{<}%
\rayz{0pt}{$\underline{\phantom{it}}$}
it is beautiful t}he
f\thinspace\graybox{4.5em}{\tt inal snuf}\thinspace{}fing o{\it f the 
lie{\bfit s of Man ALL IS LO%
\'S\rayz{4.5pt}{$\skew6\grave{}\skew6\grave{\thinspace}$}%
\rayz{6pt}{$\ddot{}$}%
\rayz{8pt}{$\acute{}$}%
\rayzb{-1.6ex}{$^7\hat{\ }$}%
\rayzb{-1.5ex}{$=$}%
T A}}{\bf LL IS L}OST
the {\it po\rlap{\ \lil /}ny he come}s 
he c\rayz{-1.9ex}{$\breve{\ }$}o\rayz{1pt}{$\underline{\phantom{co}}$}%
m\strike{5.1em}es he co{\bf mes t\kern-1pt{\bfit he} ich}or
permeat{\it es al}l MY
FAC{\it E MY FACE $^o$\kern-1pt h god n{\bfit o NO NOO\rayzb{1.2pt}{$_{_{\bowtie}}$}%
}}{\bf OO N}$\Theta$ 
stop t{\sl he an\raise 2pt\hbox{$\ast$}%
\rayz{1pt}{`}\rayz{1ex}{$\wr$}\rayzl{1ex}{$\bar{\thinspace}$}\rayzl{2.2ex}{m}%
\rayz{-1pt}{\_}\rayz{-0.5ex}{$\skew5_{{}_\vdash}$}\rayz{-2.3ex}{$\ddot{\thinspace}$}%
\rayz{-1.2ex}{\_}\rayz{-1.4ex}{\_}%
g\rayzl{-1.5ex}{w\thinspace}\rayz{1ex}{$\wr$}\rayzl{2.0ex}{m}\rayz{2.0ex}{$\neg$}%
l\rayz{2pt}{$\acute{\thinspace}$}\rayzl{1.8ex}{m}\rayzl{2.2ex}{i}\rayzl{2.6ex}{c}%
\rayz{-1.6ex}{$\hat{\thinspace}$}\rayz{-1.0ex}{$_\neg$}\rayz{-2.2ex}{$\acute{}$}%
\rayz{-1.4ex}{\_}\rayz{-1.7ex}{\_}\rayz{-1.8ex}{$\skew9_{{}_{+}}$}%
e\rayz{-2pt}{\tt \_}\rayz{-0.5pt}{$\skew6\bar{\phantom{e}}$}%
s}\graybox{2.5em}
{\tt \c{a}\rayz{-0.8ex}{$\skew8_\sqcap$}\rayzl{-1.2ex}{m}%
r\rayz{-2pt}{$^{^{\times}}$}\rayz{0pt}{$^{^{\wr}}$}\rayz{1.0ex}{$\ddot{\thinspace}$}%
\rayz{1.3ex}{$\acute{\thinspace}^{\hat{}}$}%
e} n{\bf ot
re\rayzb{0pt}{$\grave{\phantom{e}}$}\rayzl{1.5ex}{\pmb{u}}\rayzb{1ex}{$\hat{\thinspace}$}%
a\rayzl{1.2ex}{\pmb{c\thinspace}}%
\~l\rayzl{2.3ex}{\pmb{e}}\rayzb{2.2ex}{$^{^{\wr}}$}%
\rayzb{0ex}{$\skew9_{_{\vdash}}$}\rayzb{-0.4ex}{$\skew9_{_{\perp}}$}\rayzb{-1ex}{$\skew9_{_{\dashv}}$}
Z\~A\rayz{1.5ex}{/\thinspace}\rlap{\kern-2pt,}L\rayzb{0pt}{$\widetilde{\phantom{AL}}$}%
G\rlap{\kern-1.8pt'}O
IS\rayzb{-1pt}{$\check{\phantom{S}}$}\rayzb{1pt}{$\hat{\phantom{S}}$}%
\ \rayzb{-1ex}{$\frown$}\rayzb{-2ex}{''}\rayzb{-2ex}{$<$}\rayzb{-3ex}{$\dashv$}%
T\rayz{-1.5ex}{$\hat{}$\ }%
\rayz{0pt}{`\thinspace$\check{}$\ }%
\rayzl{0pt}{>\ \thinspace}%
\rayzl{3pt}{<}%
}O\rayz{-1ex}{=\thinspace}\rayz{-3pt}{,\hskip 1.5pt}\rayzb{-1ex}{$_{_{\sqcup}}$}%
\rayzb{-3.5ex}{$\grave{\thinspace}$\thinspace}%
\c{N}\rlap{\kern-2pt $^{_{\backslash}}$}\=y\rayzb{-1.5ex}{=}
TH\rayzb{-2pt}{$_{_{\dashv}}$}{\bf
\"E\rayzb{2pt}{$\acute{\phantom{E}}$}\rayzb{3pt}{$^{^{^\supset}}$}%
\rayzb{-1.6ex}{$^>\hat{\thinspace}$}
P\rayzb{0pt}{$\widetilde{\phantom{EP}}$}\rayzb{-1ex}{$\frown$}\rayzb{-1.5ex}{$\leftrightarrow$}%
\rayzb{-2.7ex}{$\hat{\hskip 0.8em}$}%
O\rayzb{1.5ex}{$\neg$}%
N\rayzb{1pt}{$\dot{\phantom{N}}$}\rayzb{0.5pt}{$^{^\smile}$}%
Y \rayzb{-1.5ex}{$\acute{\ }$\thinspace}%
H\rlap{\kern -3pt\rm /}\rayzb{-1.5ex}{$\acute{\ }$}%
\rayzb{1.7ex}{\lil c\hskip2pt}\rayzb{1.8ex}{$\sim$}\rayz{1.5ex}{$^{^{\bar{x}}}$}%
\rayzb{-1.5ex}{$\sqcap$\thinspace}\rayzb{-2.8ex}{$\hat{\phantom{H}}$}%
\c{E}\rayzb{-2.7ex}{$\check{\phantom{E}}$}\rayzb{-2.1ex}{$\imath$\thinspace}%
\rayzb{1.6ex}{\lil \"s\hskip 2pt}\rayzb{3pt}{$\acute{\phantom{E}}$}\rayzb{5pt}{$\grave{\phantom{E}}$}
\rayz{-1ex}{\bf \c{x}}\rayzb{-3.3ex}{$\hat{\ }$\hskip 2pt}\rayzb{-3.7ex}{$\hat{\ }$\hskip 2pt}%
\rayzb{1.3ex}{\lil u}\rayzb{2.0ex}{\lil s}\rayzb{2.7ex}{\lil r}%
C\rayz{0pt}{\bf -\hskip 6.5pt}\rlap{\raise 1.3pt\hbox{\kern -2pt\lil /}}%
\rayzb{1.7ex}{\lil \"t\hskip2pt}\rayzb{2.5ex}{\lil i\hskip2pt}\rayzb{3.0ex}{\lil v\hskip2pt}%
\rayzb{-2pt}{$_\vdash$}\rayzb{-1.3ex}{\lil =}\rayzb{-1.8ex}{$_\perp$}%
\rayzb{-3.9ex}{$^>\hat{\thinspace}$}%
O\rayzb{-2ex}{---\thinspace}%
\rayzb{0pt}{$\check{\phantom{O}}$}% frown sqcap perp leftrightarrow
\rayzb{-2pt}{$_\frown$}\rayzb{-5pt}{$_\sqcap$}\rayzb{-9pt}{$_\perp$}\rayzb{-12pt}{$_\leftrightarrow$}%
M\rayz{-1.1ex}{\tt -\ }%
\rayz{-1pt}{`\ }%
\rayzb{1.3ex}{\lil h\ }\rayzb{2.0ex}{\lil d\ }\rayzb{2.7ex}{\lil r\ }%
\rayzb{1.4ex}{$^{^{\neg}}$\thinspace}%
E\rayzb{-1ex}{$\smile$\thinspace}%
\rayzb{-2pt}{$_\div$}%
\rayzb{-6pt}{$_+$}%
\rayzb{-10pt}{$_\star$\hskip 1pt}%
\rayzb{-13pt}{$_{_\top}$}%
\rayzb{-1pt}{$^{^{\frown}}$}%
\rayzb{1.8ex}{\lil d\thinspace}%
\rayzb{4pt}{$^{^{\approx}}$}%
}\c{S}\rayzb{1.6ex}{$\smile$\thinspace}%
\rayzl{1.8ex}{x}\rayzl{2.2ex}{=}\rayz{5pt}{$^{^\zeta}$}%
\rayz{-1pt}{$_{_{\circ}}$}%
\rayz{-4pt}{$_{_{\varpi}}$}%
\rayz{-7pt}{$_{_{\tau}}$}%
\rayz{-11pt}{$_{_{\hat{\hskip 2pt}}}$}%
\rayz{-11pt}{$_{_{\imath}}$}%

\medskip
\parskip 10pt plus 4pt minus 4pt\parindent=0pt%
Have you tried using an XML parser instead?

\end{document}
