\def\section#1.{\bigskip{\noindent\bf#1}\par\noindent\ignorespaces}
\newcount\codenum \codenum=0
\def\code{\advance\codenum by1\item{\the\codenum.}}

\vglue 1in
\centerline{\bf SQLite Code of Conduct}
\section Overview.
The code of conduct for developers of the SQLite project is the 72
``instruments of good works'' from the 6th~century Rule of
St.~Benedict, shown below. This code of conduct has proven its mettle
in use by many thousands of communities for 1,500 years, and has
served as a baseline for many civil law codes since the time of
Charlemagne.

The SQLite developers are not monks. None are able to adhere perfectly
to this code of conduct. All will fall short. The goal of this code of
conduct is not to describe the minimum acceptable behavior below which
none transgress, but rather to describe the highest standard of
behavior to which all aspire. We grant grace to one another and
request forgiveness and promise to amend our ways when we stumble.

This code of conduct applies to the SQLite developers only. Everyone
is free to use the SQLite source code, object code, and/or
documentation without having to agree with the behavioral standards
outlined below. You are encouraged to follow these standards as in
doing so you will live a happier, healthier, and more productive
life. Nevertheless, this is a choice that you must make for yourself,
and is not a precondition for using SQLite.

\section The Code.
\code First of all, love the Lord God with your whole heart, your whole soul, and your whole strength.
\code Then, love your neighbor as yourself.
\code Do not murder.
\code Do not commit adultery.
\code Do not steal.
\code Do not covet.
\code Do not bear false witness.
\code Honor all (1 Peter 2:17).
\code Do not do to another what you would not have done to yourself.
\code Deny oneself in order to follow Christ.
\code Chastise the body.
\code Do not become attached to pleasures.
\code Love fasting.
\code Relieve the poor.
\code Clothe the naked.
\code Visit the sick.
\code Bury the dead.
\code Be a help in times of trouble.
\code Console the sorrowing.
\code Be a stranger to the world's ways.
\code Prefer nothing more than the love of Christ.
\code Do not give way to anger.
\code Do not nurse a grudge.
\code Do not entertain deceit in your heart.
\code Do not give a false peace.
\code Do not forsake charity.
\code Do not swear, for fear of perjuring yourself.
\code Utter only truth from heart and mouth.
\code Do not return evil for evil.
\code Do no wrong to anyone, and bear patiently wrongs done to yourself.
\code Love your enemies.
\code Do not curse those who curse you, but rather bless them.
\code Bear persecution for justice's sake.
\code Be not proud.
\code Be not addicted to wine.
\code Be not a great eater.
\code Be not drowsy.
\code Be not lazy.
\code Be not a grumbler.
\code Be not a detractor.
\code Put your hope in God.
\code Attribute to God, and not to self, whatever good you see in yourself.
\code Recognize always that evil is your own doing, and to impute it to yourself.
\code Fear the Day of Judgment.
\code Be in dread of hell.
\code Desire eternal life with all the passion of the spirit.
\code Keep death daily before your eyes.
\code Keep constant guard over the actions of your life.
\code Know for certain that God sees you everywhere.
\code When wrongful thoughts come into your heart, dash them against Christ immediately.
\code Disclose wrongful thoughts to your spiritual mentor.
\code Guard your tongue against evil and depraved speech.
\code Do not love much talking.
\code Speak no useless words or words that move to laughter.
\code Do not love much or boisterous laughter.
\code Listen willingly to holy reading.
\code Devote yourself frequently to prayer.
\code Daily in your prayers, with tears and sighs, confess your past sins to God, and amend them for the future.
\code Fulfill not the desires of the flesh; hate your own will.
\code Obey in all things the commands of those whom God has placed in authority over you even though they (which God forbid) should act otherwise, mindful of the Lord's precept, ``Do what they say, but not what they do.''
\code Do not wish to be called holy before one is holy; but first to be holy, that you may be truly so called.
\code Fulfill God's commandments daily in your deeds.
\code Love chastity.
\code Hate no one.
\code Be not jealous, nor harbor envy.
\code Do not love quarreling.
\code Shun arrogance.
\code Respect your seniors.
\code Love your juniors.
\code Pray for your enemies in the love of Christ.
\code Make peace with your adversary before the sun sets.
\code Never despair of God's mercy.
\bye
